Homogeneous Light Curves for Stars in Clusters from TESS

### 2019-01-30 paper/ms.tex
paper: update detrending procedure

The first is to trim and mask the images.  We convert the calibrated
image from MAST into a \texttt{fitsh}-compatible image by retrieving
the image data (omitting the uncertainty map, for now). We then trim
the image to remove virtual rows and columns (using the
\texttt{SCIROWS}, \texttt{SCIROWE}, \texttt{SCCSA}, \texttt{SCCED},
header values).  We then mask out saturated stars using the `fiign`
program with a fixed saturation level of $2^{16}\,{\rm ADU}$.  As
described by \citet{Pal_2009}, the mask is metadata to the image, and
is not actually applied to the pixel values.  During masking, we also
use the algorithm described by \citet{Pal_2009} to extend masks beyond
purely satured pixels to ``bloomed'' pixels -- pixels horizontally and
vertically adjacent to the saturated pixels.  Finally, for frames with
the \texttt{DQUALITY} bit-flag set to 32~---~corresponding to the
``momentum dumps'' described by \citet{vanderspek_2018}~---~we mask
out the entire frame.  This removes on average a few frames per
sector. Through visual inspection, we see that the stars on these
frames are extremely smeared, and are unlikely to produce useful
science data.

Next, we perform some initial analysis steps to produce metadata
needed during image subtraction.  To obtain an astrometric solution
(independent from the WCS data packaged with the frames), we use
\texttt{fistar} to perform source extraction on bright stars in each
image.  We pass the resulting source lists through
\texttt{astrometry.net} \citep{lang_2010}, which returns an
astrometric solution for each frame packaged in the WCS format
\citep[][Sec.~8]{pence_fits_2010}.  During the initial source
extraction, we also fit elongated gaussians to the bright stars,
yielding the shape parameters (S,D,K), where the flux as a function of
position is assumed to take the form
\begin{align}
  f_{\rm elong}(\vec{x}) &= B + A \exp \{ -0.5 \times ( 
    S(\Delta x^2 + \Delta y^2) + \\
    &D(\Delta x^2 - \Delta y^2) +
    K(2\Delta x \Delta y)
  )  \},
\end{align}
for $\Delta x = x-x_0$, and $\Delta y = y - y_0$.  $S$ is typically
called the ``sharpness''.  For a nearly circular shape profile, ${\rm
FWHM} \approx 2.35\sqrt{S}$ \citep[{\it e.g.},][]{Pal_2009}.  These
shape parameters are later used when selecting an astrometric
reference (\S~\ref{subsec:imagesubtraction})

@@

The second reference frame needed for image subtraction is the
photometric reference.
This is a high-SNR median average of frames that is used both to
calculate the convolution kernel during image subtraction, and to
measure the reference flux of each star.
To select viable frames for inclusion in the photometric reference, we
use the results of our initial photometry step
(\S~\ref{subsec:preparation}), and sort frames for each sector by:
\begin{enumerate}
  \item Lowest median scatter in photometry
  \item Lowest median error in photometry
  \item Lowest median background measurement
  \item Lowest median absolute deviation in background measurements
  \item Largest number of stars detected by \texttt{fiphot} with good
    flags.
    % fiphot files used for all of this.
\end{enumerate}
We convolve the 50 best candidate photometric references to the best
photometric reference, and then perform a median combination of the
frames to make the photometric reference.

@@

The preceding tedium produces lightcurves that include both
instrumental systematics as well as astrophysical variability.
We perform two subsequent detrending steps:
we decorrelate against known external parameters (EPD,
\citealt{bakos_2010}) that we know affect the stellar flux
measurements, and then we filter remaining systematic trends from
unknown parameters (TFA, \citealt{kovacs_trend_2005}).  These
procedures are the standard that have been adopted for the HAT group
(see discussions from {\it e.g.}
\citealt{huang_high-precision_2015,zhang_precision_2016}).
We briefly summarize them here for self-consistency.

In ``external parameter decorrelation'', we fit the magnitude of each star
as a function of parameters that we suspect may systematically affect
the brightness counts.
For instance, intra-pixel quantum efficiency variations can affect
the measured stellar brightness as a function of the fractional
centroid positions, $\{ x \}$ and $\{ y \}$, where $\{ x \} = x -
\lfloor x \rfloor$, for $\lfloor \cdot \rfloor$ the floor function.
Changes in the CCD electronics due to their temperature $T$ might
matter.
The fast focal ratio of the TESS cameras introduces significant
comatic aberrations to the images, which affect the shape of the point
spread function as a function of distance from the center of each
camera's field; this might be at least superficially treated by
allowing the elongated gaussian shape parameters $(S,D,K)$ to be free
parameters.

To chose among these free parameters, we plotted the magnitudes
against each parameter in the vector $\theta' = (S,D,K,S^2,K^2, D^2,
S\cdot D, S\cdot K, D\cdot K, x,y,\{x\},\{y\},T, \sin(2\pi x),
\cos(2\pi x), \sin(2\pi
y),\cos(2\pi y))$.
We used the temperature from the on-chip aluminum-copper sensor
measurements included in the engineering
data\footnote{\url{archive.stsci.edu/missions/tess/engineering/}}.

%FIXME: what did we see? What did we fit for?
We saw significant correlations for parameters X,Y, and Z.

We then fit the following model to the magnitudes $m$:
\begin{equation}
  m = {\rm const.} + \sum_i \theta_i,
\end{equation}
for 
$\theta = (S,D,K,S^2,K^2, D^2, S\cdot D, S\cdot K, D\cdot K,
x,y,\{x\},\{y\},T, \sin(2\pi x), \cos(2\pi x), \sin(2\pi y),\cos(2\pi
y))$.
We optimized via least-squares, and subtracted the maximum-likelihood
model to produce ``EPD'' lightcurves.

In ``trend filtering'', we correlate the flux timeseries of each star
against other stars in the frame (TFA, \citealt{kovacs_trend_2005}).
This requires selecting ``template stars'', which are a subsample of
star that are supposed to represent all the types of systematics
across the dataset. 
%FIXME: clarify this procedure
We select 500 template stars randomly from the stars on-chip with
$G_{\rm R_p}$ magnitudes between 8.5 and 13 (TODO: rather bright?).
We additionally require these stars to not have excessive variability,
by fitting a parabola in the RMS-magnitude plane, and discarding stars
more than $4\sigma$ away from the prediction of this fit.

>>>>

The first is to trim and mask the images.  We convert the calibrated
image from MAST into a \texttt{fitsh}-compatible image by retrieving
the image data, and omitting the uncertainty map. We then trim the
image to remove virtual rows and columns using the \texttt{SCIROWS},
\texttt{SCIROWE}, \texttt{SCCSA}, \texttt{SCCED}, header values.  We
mask out saturated stars using the `fiign` program with a fixed
saturation level of $2^{16}\,{\rm ADU}$.  As described by
\citet{Pal_2009}, the mask is metadata to the image, and is not ever
actually applied to the pixel values, except during specific image
processing steps that follow.  During masking, we also use the
algorithm described by \citet{Pal_2009} to extend masks beyond purely
saturated pixels to ``bloomed'' pixels horizontally and vertically
adjacent to the saturated pixels.  Finally, for frames with the
\texttt{DQUALITY} bit-flag set to 32~---~corresponding to the
``momentum dumps'' described by \citet{vanderspek_2018}~---~we mask
out the entire frame.  This removes on average a few frames per
sector. Through visual inspection, we see that the stars on these
frames are extremely smeared, and are unlikely to produce useful
science data.

Next, we perform some initial analysis steps to produce metadata
needed during image subtraction.  To obtain an astrometric solution
(independent from the WCS data packaged with the frames), we use
\texttt{fistar} to perform source extraction on bright stars in each
image.  We pass the resulting source lists through
\texttt{astrometry.net} \citep{lang_2010}, which returns an
astrometric solution for each frame packaged in the WCS format
\citep[][Sec.~8]{pence_fits_2010}.  During the initial source
extraction, we also fit elongated gaussians to the bright stars,
yielding the shape parameters (S,D,K), where the flux as a function of
position is assumed to take the form
\begin{align}
  f_{\rm elong}(\vec{x}) &= B + A \exp \{ -0.5 \times ( 
    S(\Delta x^2 + \Delta y^2) + \\
    &D(\Delta x^2 - \Delta y^2) +
    K(2\Delta x \Delta y)
  )  \},
\end{align}
for $\Delta x = x-x_0$, and $\Delta y = y - y_0$.  $S$ is typically
called the ``sharpness''.  For a nearly circular shape profile, ${\rm
FWHM} \approx 2.35\sqrt{S}$ \citep[{\it e.g.},][]{Pal_2009}.  These
shape parameters are later used when selecting an astrometric
reference (\S~\ref{subsec:imagesubtraction}).
We note that the fast focal ratio of the TESS cameras introduces
significant comatic aberrations to the images, stars closer to the
center of the field are more round; stars towards the edges are more
triangular.

@@

The second reference frame needed for image subtraction is the
photometric reference.  Generically, if one were to simply subtract
two arbitrary frames, the noise level in the resulting difference
image would be $\sqrt{2}$ times greater than that in the original
target frame.
To minimize this effect, we create a high-SNR median average of
$N$ selected frames, so that the noise level in the resulting
subtracted images goes as $\sqrt{1+1/N}$.
This ``photometric reference'' is used both to calculate the
convolution kernel during image subtraction, and to measure the
reference flux of each star.
To select viable frames for inclusion in the photometric reference, we
use the results of our initial photometry step
(\S~\ref{subsec:preparation}), and sort frames for each sector by:
\begin{enumerate}
  \item Lowest median scatter in photometry
  \item Lowest median error in photometry
  \item Lowest median background measurement
  \item Lowest median absolute deviation in background measurements
  \item Largest number of stars detected by \texttt{fiphot} with good
    flags.
    % fiphot files used for all of this.
\end{enumerate}
We convolve the 50 best candidate photometric references to the best
photometric reference, and then perform a median combination of the
frames to make the photometric reference.

@@

The preceding steps produce lightcurves that include both instrumental
systematics as well as astrophysical variability.  The detrending
philosophy adopted by the HAT group typically proceeeds in two
sequential stages (see discussions from {\it e.g.}
\citealt{bakos_2010,huang_high-precision_2015,zhang_precision_2016}) .
The first stage is to decorrelate against external parameters that are
known to affect the stellar flux measurements (EPD,
\citealt{bakos_2010}). The second stage is to filter remaining
systematic trends caused by unknown parameters by decorrelating the
measured brightnesses of stars against each other (TFA,
\citealt{kovacs_trend_2005}).

In general though, one must adapt detrending procedures to the data at
hand.  In \S~\ref{sec:flux_vs_external_parameters}, we show that the
``external parameter'' dependence visible in the TESS data is rather
complex: ordinary linear model-fitting, as well as an initial attempts
at non-linear model fitting, are poor descriptions of the data.  We go
on to show (\S~\ref{sec:tfa_is_good_enough}) that a plausible
detrending approach for the purpose of transit discovery is to simply
decorrelate against other nearby stars with standard TFA.

\subsubsection{Flux versus external parameters}
\label{sec:flux_vs_external_parameters}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.7\textwidth]{f3.png}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Timeseries of ``external parameters'' for a representative
    star.} {\it Top}: Instrumental raw magnitude (with a particular
    aperture size), as a function of time.  {\it Second and third from
    top}: $x$ and $y$ centroid positions as a function of time.
    Continuing in order are the CCD temperature, the $(s,d,k)$ shape
    parameters, and the measured background value.
    Most of the apparent variability is instrumental: see
    \S~\ref{sec:flux_vs_external_parameters}.
		\label{fig:external_parameter_timeseries}
	}
\end{figure*}

\begin{figure*}[t]
    \begin{center}
		\includegraphics[height=0.97\textheight]{f4.png}
    \end{center}
    \vspace{-0.8cm}
    \caption{
      {\bf Flux as a function of ``external parameters'' for four
      representative stars.} The left two columns are stars at the
      corner of a camera's field; the right two columns are from the
      centers.
      Each row shows a different parameter along the $x$-axis, given
      in text at the bottom right of each subplot.
      ``Hooks'' are common features in flux as a function of
      temperature and centroid position.
      \S~\ref{sec:flux_vs_external_parameters} gives a verbose
      description.
     \label{fig:flux_vs_external_parameters}
    }
\end{figure*}

The traditional approach to EPD 
\citep[][]{Pal_2009,bakos_2010,zhang_precision_2016} is to fit and
subtract a model of the form

\begin{equation}
  m = {\rm const.} + \sum_i c_i \theta_i,
  \label{eq:classical_EPD}
\end{equation}

where $\vec{\theta}$ is a vector of parameters such as the shape
parameters $(s,d,k)$, their products $(s^2, s\cdot d, d^2, \ldots)$,
the temperature $T$ of the CCDs\footnote{ We used the temperature from
the on-chip aluminum-copper sensor measurements included in the
engineering
data\footnote{\url{archive.stsci.edu/missions/tess/engineering/}}.  },
the centroid positions $(x,y)$, the fractional part of the
centroid positions $(\{x\},\{y\})$, and any other parameters that are
expected\footnote{ The fractional centroid positions might matter
because intra-pixel quantum efficiency variations could affect the
measured stellar brightness.  The varying temperature $T$ of the CCD
electronics might matter.  } to correlate strongly with the observed
flux.  The coefficients $c_i$ are calculated through linear
least-squares, and subtracted to produce ``EPD'' lightcurves.

The premise of this model is that the correlations between the
magnitudes and these external parameters are linear.  For ground-based
CCD data (e.g., HATNet, HATS, and Nikon DSLRs), \citet{bakos_2010} and
\citet{zhang_precision_2016} have verified that this model is a
reasonable fit to the data.

To discern whether such a model describes the TESS data, we examined scatter 
plots of each
parameter, as a function of all the other parameters.  We also
examined plots of each parameter as a function of time.
A few prominent trends were present.

\begin{enumerate}

\item {\it Flux vs. time}. Most of the lightcurves we examined showed a secular
    drift with amplitude $0.01\,{\rm mag}$ over the timescale of each
    orbit.  Sharper trends (``hooks'') at the beginning of each orbit
    seemed to be less prominent for stars at the corners of the fields
    than stars at the center.  The periodicity incurred by the
    2.5$\,{\rm day}$ momentum dumps was also noticeable in more of the
    lightcurves at the center of the field than at the corners.

\item {\it Flux vs. centroid positions}. The flux as a function of
  centroid position often showed non-linear ``hooks'' (see Figure
    \ref{fig:flux_vs_external_parameters}). Most of the data points from a 
    given orbit reside at a given
    level, but about 10\% are in a tail. This was seen in lightcurves
    all across the TESS field of view.

\item {\it Flux vs. temperature} exhibited similar hooks, with most of
  the flux values residing at a particular level, and perhaps 10\%
    following a non-linear path (often resembling the Nike ``swoosh'')
    away from the bulk of points.

\item {\it Flux vs. shape parameters}.  For lightcurves in the corner
  of the field of view, similar hooks are present in flux {\it vs.}
    $(s,d,k)$, though the hooks are less sharp.  In the center of the
    field of view, gaussian ellipses are a better description of the
    flux {\it vs.} the shape parameters.

\end{enumerate}

Considering the timeseries of parameters other than flux 
(Figure~\ref{fig:external_parameter_timeseries}):

\begin{enumerate}

\item {\it Centroid positions vs. time}.  The main variability in the
  centroid positions as a function of time is a secular drift, that is
    reset every orbit.  The 2.5 day momentum wheel dump is
    superimposed on this secular drift, and has smaller amplitude than
    the drift.

\item {\it Temperature vs. time}.  The main variability in temperature
  {\it vs.} time is a secular drift of the same timescale as that for
    the centroid positions timeseries.

\item {\it Shape parameters vs. time}.  The main variability in the
  shape parameters as a function of time is the 2.5 day momentum wheel
    dump periodicity, with hooks before each momentum dump.

\item {\it Background value vs. time}.  The background is typically stable, 
except when scattered light from the Earth or Moon enters the frame (visible 
towards the end of each orbit in 
Figure~\ref{fig:external_parameter_timeseries}).

\end{enumerate}

Given the characteristics of the variability, a linear model of the
form given in Equation~\ref{eq:classical_EPD} is not applicable.  To
fit out the correlations between flux and parameters which most
commonly exhibited ``hooks'', we explored fitting a parametric open
curve (an $N$-dimensional B-spline, \citealt{dierckx_curve_1996}) to
the flux, centroid positions, and temperatures simultaneously.  We
selected the number of knots through brute-force, by calculating
$\chi^2$ for the model fit of a grid of knots, and minimizing the
Bayesian Information Criterion.  Though this approach showed some
initial promise, even with ``optimal'' knot-selection (in the BIC
sense) it introduced undesirable residuals in the lightcurves, and
also distorted transits.

Given these complications, for the time being we omit the step of
``detrending'' as a function of external parameters. To
enable further exploration of the issue, we include all the necessary
vectors of {\it e.g.}, centroid positions, temperatures, and shape
parameters in our reported lightcurves.

\subsubsection{Trend filtering algorithm}
\label{sec:tfa_is_good_enough}

Since most of the external parameter dependence is shared between
stars, we opt to decorrelate the flux timeseries of each star
against other stars in the frame (TFA, \citealt{kovacs_trend_2005}).

%TODO: write out the procedure, give a model equation!

This requires selecting ``template stars'', which are a subsample of
star that are supposed to represent all the types of systematics
across the dataset. 
%FIXME: clarify this procedure
We select 500 template stars randomly from the stars on-chip with
$G_{\rm R_p}$ magnitudes between 8.5 and 13 (TODO: rather bright?).
We additionally require these stars to not have excessive variability,
by fitting a parabola in the RMS-magnitude plane, and discarding stars
more than $4\sigma$ away from the prediction of this fit.
